A Partitioning Algorithm for Maximum Common Subgraph Problems

### 2017-02-19 paper/paper.tex
Typo

With these definitions established, we now present \McSplit. This algorithm
finds a maximum-cardinality mapping $M^* = \{(v_1, w_1), \dots, (v_{m},
w_{m})\}$ with $|M^*| = m$ vertex pairs, where the $v_i$ are distinct vertices
from $\V(\graphG)$ and the $w_i$ are distinct vertices from $\V(\graphH)$, such
that $v_i$ and $v_j$ are adjacent in $\graphG$ if and only if $w_i$ and $w_j$
are adjacent in $\graphH$.  Given such a mapping, the subgraph of $\graphG$
induced by $\{v_i, \dots, v_{m}\}$ and the subgraph of $\graphH$ induced by
$\{w_i, \dots, w_{m}\}$ are a maximum common subgraph.

@@

The remainder of $\FuncSty{search}$ performs the search.  A label class
$\langle \setG, \setH \rangle$ is selected from $\AlgVar{future}$, where $\setG \subseteq
V(\graphG)$ and $\setH \subseteq V(\graphH)$, using some heuristic (\lineref{SelectClass}).
From this label class, a vertex $v$ is selected from $\setG$ using a heuristic
(\lineref{SelectVertex}). We now iterate over all possible vertices in $\setH$
(\linerangeref{WLoop}{ShrinkM}). A vertex $w \in \setH$ is selected and mapping $M$
is extended (\lineref{GrowM}). A new set of label-classes, $\AlgVar{future'}$,
is created (\lineref{NewFuture}).  Every label-class in $\AlgVar{future}$ can
now be split (\linerangeref{InnerLoop}{InnerLoopEnd}) into two new classes. The
first of these classes (\linerangeref{NewPWithEdge}{AddToFutureWithEdge})
contains vertices in $\setG$ adjacent to $v$ and vertices in $\setH$ adjacent to $w$.
This is added to $\AlgVar{future'}$ if both sets contain at least one vertex.
This is then repeated symmetrically for non-adjacency
(\linerangeref{NewPWithoutEdge}{InnerLoopEnd}). A recursive call is then made
(\lineref{ExpandWithV}), on return from which we remove the mapping $(v,w)$.
Having explored all possible mappings of $v$ with vertices in $\setH$ we now
consider what happens if $v$ is not matched
(\linerangeref{RemoveV}{ExpandWithoutV}).

@@

Small scale experiments (not presented here) were performed to identify
suitable heuristics for selecting a label class and a vertex within a label
class selection. Our heuristic of choice selects the label class with the
smallest $\max(|\setG|,|\setH|)$. From this class, a vertex with maximum degree
is selected from $\setG$. We discuss this further in \cref{sec:comparison}.

@@

\begin{algorithm}[t]
\DontPrintSemicolon
\nl    \For {$l \in L$}{
\nl        $\setG'' \gets \{ u \in \setG' : u \neq v \wedge A_G[v][u] = l \}$ \;
\nl        $\setH'' \gets \{ u \in \setH' : u \neq w \wedge A_H[w][u] = l \}$ \;
\nl        \If {$\setG'' \negmedspace\neq \emptyset$ \bf{and} $\setH'' \negmedspace \neq \emptyset$}{
    \nl $\AlgVar{future'} \gets \AlgVar{future'} + \langle \setG''\negmedspace, \setH'' \rangle$}
       }
    \caption{Replacement for \linerangeref{NewPWithEdge}{InnerLoopEnd} of \cref{McSplitAlg} to handle directed and labelled cases.}
\label{labDirAlg}
\end{algorithm}

@@

Our first set of experiments use a database of randomly-generated maximum
common subgraph instances
\citep{DBLP:journals/prl/SantoFSV03,DBLP:journals/jgaa/ConteFV07}.  For
unlabelled instances, we selected the first ten instances from each family
whose members have no more than 50 vertices, for a total of 4,100 instances.
For labelled instances, we selected the first ten instances from every family,
for a total of 8,140 instances with up to 100 vertices; like
\citet{DBLP:conf/cp/McCreeshNPS16}, we use the labelling scheme in which the
number of distinct vertex labels and the number of distinct edge labels is
approximately equal to 33 percent of the number of vertices in each graph.

@@

\paragraph{Unlabelled, undirected}
\cref{figure:plain-cumulative} shows a plot of cumulative runtimes
against number of problem instances solved.  We may compare
the speed of two algorithms using the horizontal distance between their curves.
For example, we could solve 2,000 of the 4,110 unlabelled undirected instances
using the \McSplit\ algorithm if a time limit of 0.5 seconds per instance were
imposed.  Its nearest competitor, CP-FC, would require a time limit of over 24
seconds per instance to solve the same number of instances.  For any given
number of instances, \McSplit\ is comfortably more than an order of magnitude
faster than its nearer competitor.  Moreover, \McSplit\ is the fastest algorithm
on 87\% of the 3,506 instances that
could be solved by at least one of the four algorithms in less than than
1,000 seconds.

@@

A further advantage of our encoding is that it gives us efficient access to a
better branching heuristic. The CP-FC algorithm uses smallest domain first,
which in our algorithm corresponds to branching on a label class with smallest
$|\setH|$. We instead branch on the label class with smallest $\min(|\setG|,|\setH|)$.
This is empirically better, and accounts for much of the difference between
the number of recursive calls made; branching on the smallest $|\setG| |\setH|$ gives
very similar results. This can be viewed as exploiting both smallest domain first,
and the dual viewpoint \citep{DBLP:conf/ecai/Geelen92} of smallest domain
first, simultaneously, but we do not have the overheads of having to maintain
and channel between the dual viewpoint that would be required when using a
conventional domain store.

>>>>

With these definitions established, we now present \McSplit. This algorithm
finds a maximum-cardinality mapping $M^* = \{(v_1, w_1), \dots, (v_{m},
w_{m})\}$ with $|M^*| = m$ vertex pairs, where the $v_i$ are distinct vertices
from $\V(\graphG)$ and the $w_i$ are distinct vertices from $\V(\graphH)$, such
that $v_i$ and $v_j$ are adjacent in $\graphG$ if and only if $w_i$ and $w_j$
are adjacent in $\graphH$.  Given such a mapping, the subgraph of $\graphG$
induced by $\{v_1, \dots, v_{m}\}$ and the subgraph of $\graphH$ induced by
$\{w_1, \dots, w_{m}\}$ are a maximum common subgraph.

@@

The remainder of the procedure performs the search.  A label class
$\langle \setG, \setH \rangle$ is selected from $\AlgVar{future}$, where $\setG
\subseteq V(\graphG)$ and $\setH \subseteq V(\graphH)$, using some heuristic
(\lineref{SelectClass}), and from this label class, a vertex $v$ is selected
from $\setG$ (\lineref{SelectVertex}). We now iterate over all 
vertices $w$ in $\setH$ , exploring the consequences of adding $(v,w)$ to $M$
(\linerangeref{WLoop}{ShrinkM}).  A new set of label-classes,
$\AlgVar{future'}$,
is created (\lineref{NewFuture}); this will be the labelling that results
from adding $(v,w)$ to our mapping.  Every label-class in $\AlgVar{future}$ can
now be split (\linerangeref{InnerLoop}{InnerLoopEnd}) into two new classes. The
first of these classes (\linerangeref{NewPWithEdge}{AddToFutureWithEdge})
contains vertices in $\setG$ adjacent to $v$ and vertices in $\setH$ adjacent to $w$.
This is added to $\AlgVar{future'}$ if both sets contain at least one vertex.
This is then repeated symmetrically for non-adjacency
(\linerangeref{NewPWithoutEdge}{InnerLoopEnd}). A recursive call is then made
(\lineref{ExpandWithV}), on return from which we remove the mapping $(v,w)$.
Having explored all possible mappings of $v$ with vertices in $\setH$ we now
consider what happens if $v$ is not matched
(\linerangeref{RemoveV}{ExpandWithoutV}).

@@

Small scale experiments (not presented here) were performed to identify
suitable heuristics for selecting a label class and a vertex within a label
class. Our heuristic of choice selects the label class with the
smallest $\max(|\setG|,|\setH|)$. From this class, a vertex with maximum degree
is selected from $\setG$. We discuss this further in \cref{sec:comparison}.

@@

\begin{algorithm}[t]
\DontPrintSemicolon
\nl    \For {$l \in L$}{
\nl        $\setG'' \gets \{ u \in \setG' : u \neq v \wedge A_G[v][u] = l \}$ \;
\nl        $\setH'' \gets \{ u \in \setH' : u \neq w \wedge A_H[w][u] = l \}$ \;
\nl        \If {$\setG'' \negmedspace\neq \emptyset$ \bf{and} $\setH'' \negmedspace \neq \emptyset$}{
    \nl $\AlgVar{future'} \gets \AlgVar{future'} \cup \{\langle \setG''\negmedspace, \setH'' \rangle\}$}
       }
    \caption{Replacement for \linerangeref{NewPWithEdge}{InnerLoopEnd} of \cref{McSplitAlg} to handle directed and edge-labelled cases.}
\label{labDirAlg}
\end{algorithm}

@@

Our first set of experiments uses a database of randomly-generated maximum
common subgraph instances
\citep{DBLP:journals/prl/SantoFSV03,DBLP:journals/jgaa/ConteFV07}.  For
unlabelled instances, we selected the first ten instances from each family
whose members have no more than 50 vertices, for a total of 4,100 instances.
For labelled instances, we selected the first ten instances from every family,
for a total of 8,140 instances with up to 100 vertices; like
\citet{DBLP:conf/cp/McCreeshNPS16}, we use the labelling scheme in which the
number of distinct vertex labels and the number of distinct edge labels is
approximately equal to 33 percent of the number of vertices in each graph.

@@

\paragraph{Unlabelled, undirected}
\cref{figure:plain-cumulative} shows a plot of cumulative 
number of instances solved against runtime.  We may compare
the speed of two algorithms using the horizontal distance between their curves.
For example, we could solve 2,000 of the 4,110 unlabelled undirected instances
using the \McSplit\ algorithm if a time limit of 0.5 seconds per instance were
imposed.  Its nearest competitor, CP-FC, would require a time limit of over 24
seconds per instance to solve the same number of instances.  For any given
number of instances, \McSplit\ is comfortably more than an order of magnitude
faster than its nearer competitor.  Moreover, \McSplit\ is the fastest algorithm
on 87\% of the 3,506 instances that
could be solved by at least one of the four algorithms in less than than
1,000 seconds.

@@

A further advantage of our encoding is that it gives us efficient access to a
better branching heuristic. The CP-FC algorithm uses smallest domain first,
which in our algorithm corresponds to branching on a label class with smallest
$|\setH|$. We instead branch on the label class with smallest $\max(|\setG|,|\setH|)$.
This is empirically better, and accounts for much of the difference between
the number of recursive calls made; branching on the smallest $|\setG| |\setH|$ gives
very similar results. This can be viewed as exploiting both smallest domain first,
and the dual viewpoint \citep{DBLP:conf/ecai/Geelen92} of smallest domain
first, simultaneously, but we do not have the overheads of having to maintain
and channel between the dual viewpoint that would be required when using a
conventional domain store.
